```{r, include=FALSE}
set.seed(42)
```

# Linear mixed models {#sec-re}

::: callout-note
#### In this chapter

In this chapter, you will learn

-   how to specify random intercepts, random slopes, nested and crossed random effects in lme4 and glmmTMB, and why REML is the default (@sec-re-specify);
-   when to model a grouping variable as a random rather than a fixed effect, and what shrinkage means (@sec-re-advantage);
-   how to check a mixed model, including convergence, the random effects and residuals per group (@sec-re-check);
-   how to get p-values, predictions, model comparisons and variance partitioning from a mixed model (@sec-re-downstream).

**Workflow steps** (@sec-intro-workflow): step 2 (design and data), step 3 (a-priori model), step 5 (residual checks), step 6 (inference).

Prerequisites: the linear model (@sec-lm), ANOVA for the part on variance partitioning (@sec-anova).
:::

## Motivation {#sec-re-motivation}

::: {#motivation}
:::

Random effects are a very common addition to regression models that are used to account for grouping (categorical) variables such as subject, year, location. To explain the motivation for these models, as well as the basic syntax, we will use an example data set containing exam scores of 4,059 students from 65 schools in Inner London. This data set is located in the R package `mlmRev`.

- Response: "normexam" (Normalized exam score).
- Predictor 1: "standLRT" (Standardized LR test score; Reading test taken when they were 11 years old).
- Predictor 2: "sex" of the student (F / M).
- Grouping factor: school

If we analyze this with a simple LM, we see that reading ability and sex have the expected effects on the exam score.

```{r chunk_chapter4_chunk24, echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
library(mlmRev)
library(effects)

mod0 = lm(normexam ~ standLRT + sex , data = Exam)
plot(allEffects(mod0))
```

However, it is reasonable to assume that not all schools are equally good, which has two possible consequences:

- If school identity correlates with sex or standLRT, school could be a confounder (more on confounding in the chapter on causal inference, @sec-causal)
- Residuals will be correlated within schools, thus they are not iid (pseudo-replication)

We could solve this problem by fitting a different mean per school

```{r}
mod0b = lm(normexam ~ standLRT + sex + school , data = Exam)
```

however, with 65 schools this would cost 64 degrees of freedom, which is a high cost just for correcting the possibility of confounding and a minor residual problem.

The solution: **Mixed / random effect models**. In a mixed model, we assume (differently to a fixed effect model) that the variance between schools originates from a normal distribution. There are different interpretations of a random effect (more on this in @sec-re-interpretation), but one interpretation is to view the random effect as an additional error, so that such a mixed model has two levels of error:

- First, the random effect, which is a normal "error" acting on an entire group of data points (in this case school).

- And second, the residual error, which is a normal error per observation and acts on top of the random effect

Because of this hierarchical structure, these models are also called "multi-level models" or "hierarchical models".

::: callout-note
Because grouping naturally occurs in any type of experimental data (batches, blocks, observer, subject etc.), random and mixed effect models are the de-facto default for most experimental data! Naming conventions:

- No random effect = fixed effect model
- Only random effects = random effect model
- Random effects + fixed effects = mixed model

Most models that are used in practice are mixed models
:::

## Specifying and estimating mixed models {#sec-re-specify}

::: {#specifying-and-estimating-mixed-models}
:::

### The random intercept model {#sec-re-intercept}

The simplest random effect structure is the random intercept model.

Each random effect model has a fixed counterpart. For the random intercept model, this counterpart is the model we already saw in @sec-re-motivation, with school as a fixed effect

```{r}
mod0b = lm(normexam ~ standLRT + sex + school , data = Exam)
```

The random intercept model is practically identical to the model above, except that it assumes that the school effects come from a common normal distribution. The model can be estimated with `lme4::lmer`

```{r chunk_chapter4_chunk25, echo=TRUE, eval=TRUE}
library(lme4)
mod1 = lmer(normexam ~ standLRT + sex +  (1 | school), data = Exam)
summary(mod1)
```

If we look at the outputs, you will probably note that the `summary()` function doesn't return p-values. More on this in @sec-re-df. Also, we see that the effects of the individual school are not explicitly mentioned. The summary only returns the mean effects over all schools, and the variance between schools. However, you can obtain the individual school effects via

```{r, eval=F}
ranef(mod1)
```

Moreover, unlike the outputs we have seen before, the model also returns a correlation between fixed effect estimates. This, however, is simply a choice of the programmers, it has nothing to do with the random effect models - you could calculate the same output for the `lm()` via the function `vcov()`.

Here is a visualization of the fitted effects for standLRT

```{r chunk_chapter4_chunk26, echo=F, eval=TRUE}
with(Exam, {
  randcoef = ranef(mod1)$school[,1]
  fixedcoef = fixef(mod1)
  plot(standLRT, normexam)
    for(i in 1:65){
      abline(a = fixedcoef[1] + randcoef[i], b = fixedcoef[2], col = i)
    }
})
```

We can see that there is the same slope, but a different intercept per school.

The RE effectively "centers" the categorical predictor - unlike for the fixed effect model, where the intercept would be interpreted as the value for the first school, the intercept in the random effect model is the mean across all schools, and the REs measure the deviation of the individual school from the mean. It should be stressed, however, that this is a welcome consequence, but not the main motivation, for the RE structure.

::: {.callout-tip collapse="true"}
#### Generate equation with the `equatiomatic` package

We can use the `equatiomatic` package to print LaTeX code for the underlying equation of our random intercept model:

```{r, eval=FALSE}
library(equatiomatic)
extract_eq(mod1)
```

$$\begin{aligned}
  \operatorname{normexam}_{i}  &\sim N \left(\alpha_{j[i]} + \beta_{1}(\operatorname{standLRT}) + \beta_{2}(\operatorname{sex}_{\operatorname{M}}), \sigma^2 \right) \\
    \alpha_{j}  &\sim N \left(\mu_{\alpha_{j}}, \sigma^2_{\alpha_{j}} \right)
    \text{, for school j = 1,} \dots \text{,J}
\end{aligned}$$
:::

::: {#tip-re-design .callout-tip}
#### From study design to random effects

The random effect structure should follow from the design of the study, not from the data. Ask yourself:

-   **What is the experimental unit**, i.e. the unit to which a treatment was applied, or at which a predictor varies? If you have several measurements per unit (several plants per plot, several visits per site), these measurements are not independent replicates. Treating them as if they were is called **pseudoreplication** [@hurlbert1984pseudoreplication] and makes p-values too small. A random intercept for the unit (here `(1|plot)`) accounts for this when the treatment varies between units: the effective sample size is then roughly the number of plots, not the number of plants. The same applies to any predictor that is constant within a unit, e.g. the elevation of a site that was surveyed repeatedly.
-   **How are the units grouped?** Typical designs and their random effects:
    -   randomized blocks: `y ~ treatment + (1|block)`
    -   several measurements per plot (subsampling): `y ~ treatment + (1|plot)`
    -   split-plot (factor A on main plots, factor B on subplots within the main plots): `y ~ A * B + (1|block/mainplot)` (`/` means nested: main plots within blocks, see @tip-re-nested)
    -   repeated measurements of the same units over time: `y ~ treatment * time + (1|site)`, possibly with a temporal correlation structure (@sec-corr-temporal)
    -   two independent grouping factors, e.g. sites and observers: `y ~ x + (1|site) + (1|observer)` (crossed random effects, @tip-re-nested)
-   **Does the grouping factor have enough levels?** The variance of a random effect is estimated from its levels. A common rule of thumb is therefore that a random effect needs at least about 5-10 levels, and that a grouping variable with fewer levels is better modeled as a fixed effect [@bolker2015linear; @harrison2018brief]. The discussion is, however, nuanced (see @sec-re-usage): with few levels, singular fits (variance estimated as zero) are common, but simulations suggest that the tests of the population-level effects can still be reliable [@oberpriller2022fixed].
-   **Is there structure left after that?** Correlation within units over time (@sec-corr-temporal) or changing variance (@sec-disp) can be modeled in addition. The college student case study (@sec-re-college) shows a worked example.
:::

### The random slope model {#sec-re-slope}

A second effect that we could imagine is that the effect of `standLRT`{.R} is different for each school. As a fixed effect model, we would express this, for example, as `lm(normexam ~ standLRT*school + sex)`{.R}

The random effect equivalent is called the random slope model. A random slope model assumes that each school also gets their own slope for a given parameter (per default we will always estimate slope and intercept, but you could overwrite this, not recommended!). Let's do this for standLRT (you could of course fit random slopes on both predictors as well).

```{r chunk_chapter4_chunk27, echo=TRUE, eval=TRUE}
mod2 = lmer(normexam ~ standLRT + sex +  (standLRT | school), data = Exam)
summary(mod2)
```

Again, the difference to the fixed effect model is that in the random effect model, we assume that the variance between slopes comes from a normal distribution. The result is similar to the random intercept model, except that we have an additional variance term for the slopes, and a term for the interaction between slopes and intercepts.

::: {.callout-tip collapse="true"}
#### Generate equation with the `equatiomatic` package

We can use the `equatiomatic` package to print LaTeX code for the underlying equation of our random slope model:

```{r, eval=FALSE}
library(equatiomatic)
extract_eq(mod2)
```

$$\begin{aligned}
  \operatorname{normexam}_{i}  &\sim N \left(\alpha_{j[i]} + \beta_{1j[i]}(\operatorname{standLRT}) + \beta_{2}(\operatorname{sex}_{\operatorname{M}}), \sigma^2 \right) \\    
\left(
  \begin{array}{c} 
    \begin{aligned}
      &\alpha_{j} \\
      &\beta_{1j}
    \end{aligned}
  \end{array}
\right)
  &\sim N \left(
\left(
  \begin{array}{c} 
    \begin{aligned}
      &\mu_{\alpha_{j}} \\
      &\mu_{\beta_{1j}}
    \end{aligned}
  \end{array}
\right)
, 
\left(
  \begin{array}{cc}
     \sigma^2_{\alpha_{j}} & \rho_{\alpha_{j}\beta_{1j}} \\ 
     \rho_{\beta_{1j}\alpha_{j}} & \sigma^2_{\beta_{1j}}
  \end{array}
\right)
 \right)
    \text{, for school j = 1,} \dots \text{,J}
\end{aligned}$$
:::

Here a visualization of the results

```{r chunk_chapter4_chunk28, echo=TRUE, eval=TRUE}
with(Exam, {
  randcoefI = ranef(mod2)$school[,1]
  randcoefS = ranef(mod2)$school[,2]
  fixedcoef = fixef(mod2)
  plot(standLRT, normexam)
    for(i in 1:65){
      abline(a = fixedcoef[1] + randcoefI[i] , b = fixedcoef[2] + randcoefS[i], col = i)
    }
})
```

### Mixed models in different packages {#sec-re-package}

#### lme4 / glmmTMB {#sec-re-lme4}

There are a large number of packages to fit mixed models in R. The main packages that you use, because they are most user-friendly and stable, are lme4 and glmmTMB. They share more or less the same syntax for random effects.

- Random intercept: `(1 | group)`.
- ONLY random slope for a given fixed effect: `(0 + fixedEffect | group)`{.R}.
- Random slope + intercept + correlation (default): `(fixedEffect | group)`{.R}.
- Random slope + intercept without correlation: `(fixedEffect || group)`{.R}, identical to\
  `(1 | group) + (0 + fixedEffect | group)`{.R}.
- Nested random effects: `(1 | group / subgroup)`{.R}. If groups are labeled A, B, C, ... and subgroups 1, 2, 3, ..., this will create labels A1, A2 ,B1, B2, so that you effectively group in subgroups. Useful for the many experimental people that do not label subgroups uniquely, but otherwise no statistical difference to a normal (crossed) random effect.
- Crossed random effects: You can add random effects independently, as in\
  `(1 | group1) + (1 | group2)`{.R}.

::: {#tip-re-nested .callout-tip collapse="true"}
#### More on crossed vs. nested REs

There is often confusion about the behavior and meaning of crossed vs. nested REs. From the statistical / mathematical side, there is no difference between them, there are just REs. Crossed vs. nested is a shorthand to tell you how the data was coded.

Let's look at the Exam data, for example. If you run

```{r, eval = F}
table(Exam$student, Exam$school)
```

you will see that the student ID = 1 appears in many schools. Clearly, this cannot be the same student, the reason is that each school uses their own ID. If you would fit a crossed RE, as in

```{r, eval = F}
m1 = lmer(normexam ~ standLRT + sex +  (1 | school) + (1 |student), data = Exam)
summary(m1)
```

you would fit a common effect for each student with ID = 1. This, however, doesn't make sense, there is no connection between these students. Thus, you should fit

```{r, eval = F}
m2 = lmer(normexam ~ standLRT + sex +  (1 | school/student), data = Exam)
summary(m2)
```

Compare the number of RE groups in the summary: in m1, you have 650 students. In m2, you have 4055 student:school levels for the RE. Thus, the nested syntax makes sure each individual student gets his/her own estimated value.

You could fit the same mod by changing the coding of student. To do this, we create a new variable, combining school and student into a unique ID

```{r, eval= F}
Exam$uniqueStudent = as.factor(paste(Exam$school, "-", Exam$student))

m3 = lmer(normexam ~ standLRT + sex +  (1 | school) + (1|uniqueStudent), data = Exam)
summary(m3)

```

If you check the summary of m3, now we also have 4055 levels for uniqueStudent, and the results are identical to the nested model version.
:::

#### nlme {#sec-re-nlme}

nlme is a classic package for fitting (nonlinear) mixed effect models. I would argue that it is largely superseded by lme4 and glmmTMB, but it still offers a few special tricks that are not available in either lme4 or glmmTMB. A linear mixed model with random intercept in nlme looks like this

```{r, eval = F}
nlme::lme(Resp ~ Predictor, random = ~1|Group, data = data)
```

#### mgcv {#sec-re-mgcv}

If you want to use GAMs, there is basically only mgcv (or brms, if you're willing to go Bayesian).

```{r, eval = F}
# Group must be a factor
mgcv::gam(Resp ~ Predictor + s(Group, bs = 're'), data = data)
```

#### brms {#sec-re-brms}

brms is a very interesting package that allows you to estimate mixed models via Bayesian Inference with STAN.

```{r, eval = F}
brm_mod <- brms::brm(Resp ~ Predictor + (1|Group),
                     data = data) # add e.g. family = poisson for GLMMs, see the chapter on GLMs
summary(brm_mod)
```

A complete brms GLMM is shown in the owls exercise (@cau-glm-owls). For an introduction to Bayesian mixed models, see [the chapter on linear and linear mixed models](https://florianhartig.github.io/LearningBayes/2C-LMM.html) in @hartig2024introduction.

### REML vs. MLE {#sec-re-reml}

Frequentist linear mixed models are usually estimated with REML = restricted maximum likelihood, rather than with classical MLE. Most packages allow you to switch between REML and MLE. The motivation for REML is that for limited data (non-asymptotics), the MLE for variance components of the model are typically negatively biased, because the uncertainty of the fixed effects is not properly accounted for. If you go through the mathematics, you can make an analogy to the bias-corrected sampling variance [@bates2011computational].

REML maximizes the likelihood of linear combinations of the data that do not depend on the fixed effects (equivalently, the likelihood after integrating out the fixed effects). This accounts for the df used by the fixed effects, like dividing by n-p instead of n when estimating a variance in a linear model.

Most packages offer an option to switch between REML and ML estimation of the model. In lme4, this is done via

```{r, eval = F}
m = lmer(normexam ~ standLRT + sex +  (1 | school), data = Exam)
m = lmer(normexam ~ standLRT + sex +  (1 | school), data = Exam, REML = F)
```

Regarding when to use what, use the following rule of thumb:

- Per default, use REML

- Switch to ML if you use downstream any model functions that use the likelihood, unless you are really sure that the calculation you do is also valid for REML.

::: callout-tip
Later in this chapter (@sec-re-selection), we will talk about model selection via LRTs and AIC. Because those depend on the likelihood, it is generally recommended to switch from REML to ML in this case. The reason is that, because REML estimates the REs conditional on the fixed effects, REML likelihoods are not directly comparable if fixed effects are changed. I would recommend sticking to this, although I have to say that I have always wondered how big this problem is, compared to the other df problems associated to mixed models (see @sec-re-df).
:::

::: {#cau-re-plantheight .callout-caution icon="false"}
#### Exercise — A mixed model for plantHeight

Take the plantHeight model that we worked with already in the exercise in the chapter on linear regression (@cau-lm-plantheight):

```{r, eval=FALSE}
library(EcoData)

oldModel <- lm(loght ~ temp, data = plantHeight)
summary(oldModel)
```

We have one observation per species. Consider that the relationship height \~ temp may be different for each family. Some families may have larger plants in general (random intercept), but it may also be that the temperature relationship changes per family (random slope). Run these two models for our problem.
:::

::: {.callout-tip collapse="true" appearance="minimal" icon="false"}
#### Solution

```{r, eval = F}
# Random intercept model

library(lme4)
library(lmerTest)
randomInterceptModel <- lmer(loght ~ temp + (1|Family), data = plantHeight)
summary(randomInterceptModel)

# Random slope model

randomSlopeModel <- lmer(loght ~ temp + (temp | Family), data = plantHeight)
summary(randomSlopeModel)

# scaling helps the optimizer
plantHeight$sTemp = scale(plantHeight$temp)

randomSlopeModel <- lmer(loght ~ sTemp + (sTemp | Family), data = plantHeight)
summary(randomSlopeModel)
```
:::

## Advantages and interpretation of random effects {#sec-re-advantage}

::: {#advantages-and-interpretation-of-random-effects}
:::

What do the random effects mean, and why should we use them?

### Interpretation {#sec-re-interpretation}

The classical interpretation of the random intercept is that they model a group-structured residual error, i.e. they are like residuals, but for an entire group. However, this view breaks down a bit for a random slope model.

Another view this is that REs terms absorb group-structured variation in model parameters. This view works for random intercept and slope models. So, our main model fits the grand mean, and the RE models variation in the parameters, which is assumed to be normally distributed.

A third view of RE models is that they are regularized fixed effect models. What does that mean? Note again that, conceptually, every random effect model corresponds to a fixed effect model. For a random intercept model, this would be

```{r, eval = F}
lmer(y ~ x + (1|group)) => lm(y ~ x + group)
```

and for the random slope, it would be

```{r, eval = F}
lmer(y ~ x + (x|group)) => lm(y ~ x * group)
```

So, alternatively to a mixed model, we could always fit a fixed effect model, and that would also take care of the variation in groups. Of course, the output of the two models differs, but that's just what people programmed. The real difference, however, is that by making the assumption that the random effects come from a normal distribution, we are imposing a constraint on the estimates, which creates a shrinkage on the fitted REs (see @sec-re-shrinkage).

### RE shrinkage {#sec-re-shrinkage}

To understand shrinkage, note that whatever the fitted sd of the normal distribution underlying the RE, as long as it is finite, it will always be more likely that the RE estimate is at zero (i.e. the grand mean) than at 1 or 2 sd. This means that RE estimates are biased towards the grand mean.

We can visualize the effect by comparing estimates of a fixed to a random effect model

```{r}
mf = lm(normexam ~ 0 + school, data = Exam)
mr = lmer(normexam ~ (1 | school), data = Exam)

ef = coef(mf)
er = ranef(mr)$school$`(Intercept)`
plot(ef, er, xlab = "fixed", ylab = "random")
abline(0,1)
```

The plot shows that random effect estimates are biased towards the grand mean, i.e. positive values are lower than in the LM, and negative values are higher than in the LM. This is called shrinkage.

::: callout-note
The shrinkage imposed by the RE is equivalent to an L2 (ridge) regression (see @sec-select-shrinkage), or a normally distributed Bayesian prior. You can see this by considering that the log likelihood of a normal distribution is equivalent to a quadratic penalty for deviations from the mean. The difference is that the strength of the penalty, controlled by the RE sd, is estimated. For that reason, the RE shrinkage is also referred to as an example of "adaptive shrinkage".
:::

This RE shrinkage is useful in many situations, beyond describing a group-structured error in the data. A common use is that by using a random slope model, parameter estimates for groups with few data points are drawn towards the grand mean and thus informed by the estimates of the other groups. For example, if we have data for common and rare species, and we are interested in the density dependence of the species, we could fit a random slope for density per species. Consider the plantHeight case study we did in the exercise in the chapter on linear regression (@cau-lm-plantheight-revisited). We saw that, even after re-leveling the predictor, not all interactions could be estimated, because some factor levels had only one observation.

```{r, results='hide'}
library(EcoData)
plantHeight$growthform2 = relevel(as.factor(plantHeight$growthform), "Herb")
plantHeight$sTemp = scale(plantHeight$temp)
fit <- lm(loght ~ sTemp * growthform2 , data = plantHeight)
summary(fit)
```

If you run the same thing as a random slope model, you will see that we can estimate a random intercept and slope for each species, even those that have only one observation

```{r}
fit <- lmer(loght ~ sTemp + (sTemp | growthform2) , data = plantHeight)
ranef(fit)
```

The reason is that rare growth forms will be constrained by the RE, while common growth forms with a lot of data can overrule the normal distribution imposed by the random slope and get their own estimate. In this picture, the random effect imposes an adaptive shrinkage, similar to a ridge shrinkage estimator, with the shrinkage strength controlled by the standard deviation of the random effect. An alternative view is that the RE variance acts as a prior distribution on the slope. Of course, estimates are biased towards zero, so if you want to interpret these estimates that is a cost that you have to take into account.

### When to use what? {#sec-re-usage}

So, to finalize this discussion: there are essentially two main differences between a random vs. fixed effect model

1.  In the RE model, we are fitting a grand mean and differences from the grand mean, while in the fixed effect model, we have to set up contrasts (various options, see @sec-lm-contrast), with treatment contrasts as defaults
2.  The LM (OLS) is mathematically proven to be the best linear unbiased estimator (BLUE) of the regression problem. In a mixed model, REs (but not fixed effects) are biased towards the grand mean, but the advantage is that the random effect loses fewer degrees of freedom, i.e. the power on the fixed effects will be higher.

While point 1 is more a choice about how to report results (in principle, we could also calculate the grand mean for fixed effect group estimates), the second point (higher power) is a clear advantage in many cases, in particular if you don't care about the bias on the REs. Therefore, the rule of thumb for REs is:

- If you have a categorical variable, and you're not interested in its estimates, model it as a RE

- If you want to get unbiased estimates of the categorical variable (or its interactions), use a fixed effect

A common question is how many levels a grouping variable needs to be modeled as a random effect. Because the variance of a random effect is estimated from its levels, a common rule of thumb is to use at least about 5-10 levels, and to model a grouping variable with fewer levels as a fixed effect [@bolker2015linear; @harrison2018brief]. With few levels, the variance estimate is indeed imprecise and biased towards zero, and singular fits are common. However, if you are interested in the population-level effect of a predictor rather than in the individual levels, simulations by @oberpriller2022fixed suggest that a mixed model can still give reliable tests, provided that singular fits are handled. They recommend to start with a mixed model regardless of the number of levels. If the effect of the predictor varies between groups, start with a correlated random slope and intercept; after a singular fit, remove the correlation (`||`, see @sec-re-convergence), and if the fit is still singular, switch to a fixed effect. In their simulations, singular mixed models had inflated type I error rates, similar to a model that omits the grouping variable, and omitting random slopes that are present in the data also inflated the type I error. In short, with few levels both options can be defended, and more levels are clearly better, so when planning a study, aim for more than the minimum.

The choice between a random and a fixed effect based on whether you are interested in the estimates is just a rule of thumb; in data-poor situations, it is common to use random slope models and interpret the estimates as well. In this case, we are using the RE to optimize the bias-variance trade-off (see the section on the bias-variance trade-off in the model selection chapter, @sec-select-biasvar).

## Model checks {#sec-re-check}

::: {#model-checks}
:::

As residual checks for an LMM, you should first of all check for everything that you would check in an LM, i.e. you would expect that the residual distribution is iid normal. On top of that, you should check the REs for:

1.  Convergence (mixed models often have convergence problems)
2.  Normality
3.  Correlation with predictors -\> RE can mask misfit
4.  Residual patterns -\> evidence for random slope

I will show all of those using the example

```{r}
mod1 = lmer(normexam ~ standLRT + sex +  (1 | school), data = Exam)
summary(mod1)
```

### Convergence {#sec-re-convergence}

If there is a convergence problem, you should usually get a warning. Other signs of convergence problems may be RE estimates that are zero, CIs that are extremely large, or strong correlations in the correlation matrix of the predictors that is reported by `summary()`. Here, we have no problems, but we'll see examples of this as we go on.

There are different types of convergence problems that you can encounter:

- Warnings about scaling and centering -\> just do it, it doesn't cost you!
- The warning "boundary (singular) fit: see help('isSingular')" (meaning that some dimensions of the variance-covariance matrix have been estimated to zero. -\> often, this just means that a RE estimate is zero. In many cases, this is not a problem and can actually be ignored, but it is hard to give general advice to when this is the case. With few levels in the grouping variable, however, a singular fit can inflate the type I error of the fixed effects [@oberpriller2022fixed]; see the recommendations in @sec-re-usage. In random slope models, it often helps to move from \| to \|\| (suppressing the correlation). Removing the RE will solve the problem, but could lead to type I error inflation. A more conservative solution would be to change the RE to a fixed effect
- Optimizer / convergence warnings: use `lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e5))` (`glmerControl()` for `glmer`) to change the optimizer to bobyqa and possibly increase the number of iterations, e.g. `lmer(..., control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e5)))`.

### Observation-level residual plots {#sec-re-residual}

Unfortunately, lme4 plots only provide res \~ fitted as default

```{r}
plot(mod1)
```

For a qqplot, you have to calculate the residuals by hand

```{r}
qqnorm(residuals(mod1))
qqline(residuals(mod1))
```

Of course, any time you do a qqnorm plot, you could also run a shapiro.test to get a p-value for the deviation from normality (interpret visually: such tests are over-sensitive for large n and underpowered for few groups). You can also use the DHARMa plots

```{r, message=FALSE}
library(DHARMa)
res <- simulateResiduals(mod1, plot = T)
```

Both plots look fine. Of course, we should also check residuals against predictors. In DHARMa, we can do this via

```{r}
par(mfrow = c(1,2))
plotResiduals(res, Exam$standLRT)
plotResiduals(res, Exam$sex)
```

::: {#tip-re-predterms .callout-tip}
#### Conditional, unconditional and marginal predictions

For mixed models, there are three ways to make predictions, which are easy to confuse:

-   **Conditional** predictions include the estimated random effects, i.e. they are the predictions for the particular groups in the data (`re.form = NULL` in lme4, the default).
-   **Unconditional** predictions use only the fixed effects, i.e. the random effects are set to zero (`re.form = NA` or `re.form = ~0`). This is the prediction for an "average" group.
-   **Marginal** predictions are averaged over the distribution of the random effects, i.e. the mean prediction over all possible groups.

In a linear mixed model, unconditional and marginal predictions are identical, because the random effects average to zero. With a nonlinear link function (GLMMs), they differ (see @sec-glm-marginal). Note that these terms are not used consistently in the literature and in R packages; e.g. the "marginal R^2^" of mixed models (@sec-re-anova) is based on the fixed effects only.
:::

::: {#tip-re-predtype .callout-tip}
#### Conditional vs. unconditional predictions and residuals

When you predict with a random effect model, you can do conditional or unconditional predictions (see @tip-re-predterms). The difference is that conditional predictions include the REs, while unconditional predictions only include the fixed effects.

```{r, eval = F}
predict(mod1, re.form = NULL) # conditional, default
predict(mod1, re.form = ~0) # unconditional
```

This also makes a difference for the residual calculations: in principle, you could calculate residuals with respect to both predictions. In lme4, however, `residuals()` always returns conditional residuals (observed minus conditional predictions). It has no `re.form` argument, and additional arguments are silently ignored, so `residuals(mod1, re.form = ~0)` would still give conditional residuals. Unconditional residuals have to be calculated by hand with `predict()`:

```{r, eval = F}
residuals(mod1) # conditional
Exam$normexam - predict(mod1, re.form = NULL) # conditional, same as residuals(mod1)
Exam$normexam - predict(mod1, re.form = ~0) # unconditional
```

DHARMa offers both: since version 0.5.0, `simulateResiduals()` simulates conditional on the fitted random effects by default, and with `simulateREs = "unconditional"` it re-simulates them. Both have advantages and disadvantages:

1.  Conditional predictions are more sensitive to violations of the iid normal assumptions of the outer (observation-level) normal error
2.  However, REs can sometimes absorb misfit, such that plotting conditional residuals \~ pred looks fine, but unconditional \~ pred does not. For that reason, if you use unconditional predictions, you should also check REs \~ pred (see @sec-re-predictor)
:::

### Normality of REs {#sec-re-normality}

I prefer to look at this visually. You can also run a shapiro.test() to check if you should consider the pattern at all, but interpret it visually: such tests are over-sensitive for large n and underpowered for few groups.

```{r}
x = ranef(mod1)
qqnorm(x$school$`(Intercept)`)
```

Question is how bad can this be? [@neuhaus2013] report some simulation results that suggest that fixed effect slopes (bias, coverage) are relatively robust to misspecification in the RE distribution, which is reassuring.

### Correlation of REs with predictors {#sec-re-predictor}

As mentioned in @tip-re-predtype, REs can sometimes absorb misfit of the fixed effect structure. It can therefore be useful to check if there is a correlation with the predictors, especially if you only check conditional residuals. Example:

```{r}
y = aggregate(cbind(standLRT, sex) ~ school, FUN = mean, data = Exam)
plot(x$school$`(Intercept)` ~ y$standLRT)
```

### Residual pattern per group {#sec-re-pergroup}

lme4 has an excellent plot syntax to plot residuals per grouping factor. Check out the help of `plot.merMod`. The following plot shows residuals \~ fitted for each school

```{r}
plot(mod1, resid(., scaled=TRUE) ~ standLRT | school, abline = 0)
```

Looks still good to me, but if we think we see a pattern here, we should introduce a random slope.

## Downstream analysis of mixed models {#sec-re-downstream}

::: {#downstream-analysis-of-mixed-models}
:::

Specifying mixed models is quite simple, but their practical application raises several issues (degrees of freedom, model selection, ANOVA, predictions). The following table gives an overview of how to obtain tests and confidence intervals for mixed models; the subsections of this section explain the details. The columns for `glmer` and `glmmTMB` become relevant for generalized linear mixed models (@sec-glm-glmm).

| Question | `lmer` | `glmer` | `glmmTMB` | Section |
|----|----|----|----|----|
| Test a single fixed effect | t-test with approximate df (`summary()` after loading `lmerTest`) | Wald z-test (`summary()`) | Wald z-test (`summary()`) | @sec-re-df |
| Test a term with several df (e.g. a factor) | F-test with approximate df (`anova()` after loading `lmerTest`, type III); alternatives in @sec-re-anova | `car::Anova()` (Wald $\chi^2$) | `car::Anova()` (Wald $\chi^2$) | @sec-re-anova |
| Compare groups or treatments (contrasts) | `emmeans()` with `pairs()` or `contrast()` | `emmeans()` | `emmeans()` | @nte-anova-emmeans |
| Compare nested models with different fixed effects | likelihood ratio test, `anova(m0, m1)` (refits REML models with ML) | `anova(m0, m1)` | `anova(m0, m1)` | @sec-re-fixed |
| Test a random effect | simulated LRT (`DHARMa::simulateLRT()`); `lmerTest::ranova()` is approximate and conservative | simulated LRT | simulated LRT | @sec-re-simlrt |
| CIs for fixed effects | `confint()` (profile); for few groups (fewer than about 10-20), parametric bootstrap (`confint()` of a `bootMer()` result) | `confint()` (profile; `method = "Wald"` is faster) | `confint()` (Wald; `method = "profile"` is also available) | @sec-re-df |
| CIs for random effect standard deviations | `confint()` (profile) | `confint()` (profile) | `confint()` (Wald, listed as `Std.Dev.`) | @sec-re-df |
| CIs for predictions | parametric bootstrap with `bootMer()` (the prediction function also needs `re.form = NA`); approximate: `predict(..., se.fit = TRUE)` (lme4 2.0 or later) | as for `lmer` | `predict(..., se.fit = TRUE, re.form = NA)`; compute the CI on the link scale and back-transform | @sec-re-prediction |

: Tests and confidence intervals for mixed models. {#tbl-re-inference}

All of these are approximations, because the degrees of freedom of a mixed model are not clearly defined (@sec-re-df). With few groups (as a rule of thumb, fewer than about 10-20), Wald tests and also likelihood ratio tests with the standard $\chi^2$ distribution give p-values for fixed effects that are too small. For linear mixed models, the approximate degrees of freedom of `lmerTest` (Satterthwaite) or Kenward-Roger (`car::Anova(m, test.statistic = "F")`) are then more reliable; for GLMMs, use a parametric bootstrap or a simulated LRT (`DHARMa::simulateLRT()`, `pbkrtest::PBmodcomp()`), and profile instead of Wald CIs. In contrast, a standard likelihood ratio test for a random effect is conservative, because the null hypothesis (variance = 0) lies at the boundary of the parameter space, which is another reason to use the simulated LRT. Finally, note that `anova()` after loading `lmerTest` gives type III tests, while `car::Anova()` gives type II tests by default; with interactions, they test different hypotheses (@sec-anova-collinearity, @cau-re-typeiii), so say which one you report.

### Degrees of freedom for a random effect {#sec-re-df}

Possibly the most central problem is: How many parameters does a random effect model have?

::: callout-note
Why is it so important to know how many parameters a model has? The answer is that what seems like a very minor point is needed in all the mathematics for calculating p-values, AIC, LRTs and all that, so without knowing the complexity of the model, the mathematics that is used in `lm()` breaks down.
:::

We can make some guesses about the complexity of an LMM by looking at its fixed effect version counterparts:

```{r eval=F}
mod1 = lm(normexam ~ standLRT + sex , data = Exam)
mod1$rank # 3 parameters.

mod2 = lmer(normexam ~ standLRT + sex +  (1 | school), data = Exam)
# No idea how many parameters.

mod3 = lm(normexam ~ standLRT + sex + school, data = Exam)
mod3$rank # 67 parameters.
```

So, we see that the fixed effect model without school has 3 parameters, and with school 67. It is reasonable to assume that the complexity of the mixed model is somewhere in-between, because it fits the same effects as mod3, but they are constrained by the normal distribution.

The width of this normal distribution is controlled by the estimated variance of the random effect. For a high variance, the normal is very wide, and the mixed model is nearly as complex as mod3; but for a low variance, the normal is very narrow, and the mixed model it is only as complex as mod1.

::: callout-note
Technically, the mixed model mod2 actually has one parameter more than the fixed effect model mod3 - it also estimates the standard deviation of the random effects. Could it thus ever be more flexible than mod3? The answer is no - the RE standard deviation is what is technically called a hyperparameter. It is not an effect that is estimated, but rather a parameter that controls the freedom of the estimated effects. The number of parameters that are used to estimate the response are identical in mod2 and mod3.
:::

Because of these issues, `lmer`{.R} by default does not return p-values (for an overview of the options, see @tbl-re-inference). The help advises you to use bootstrapping to generate valid confidence intervals and p-values on parameters. This is possible indeed, but very cumbersome.

However, you can calculate p-values based on approximate degrees of freedom via the `lmerTest`{.R} package, which also corrects ANOVA for random effects, but not AIC.

```{r chunk_chapter4_chunk32, echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
library(lmerTest)

m2 = lmer(normexam ~ standLRT + sex +  (1 | school), data = Exam)
summary(m2)
```

Confidence intervals for the fixed effects and the random effect standard deviations (`.sig01` is the standard deviation between schools, `.sigma` the residual standard deviation) can be calculated by profiling the likelihood:

```{r, message=FALSE}
confint(m2)
```

(Refit with `REML = FALSE` only if you want to run LRTs or AIC comparisons on fixed effects.)

::: {.callout-caution icon="false"}
#### Exercise — Mixed vs. fixed effect models

For the plantHeight mixed models in the exercise above (@cau-re-plantheight), compare the p-values of the slope for temp to the fixed-effect equivalent (adding family as main effect / interaction). What do you find?
:::

::: {.callout-tip collapse="true" appearance="minimal" icon="false"}
#### Solution

```{r, eval = F}
fixedInterceptModel <- lm(loght ~ sTemp + Family, data = plantHeight)
randomInterceptModel <- lmer(loght ~ sTemp + (1 | Family), data = plantHeight)
summary(fixedInterceptModel)
summary(randomInterceptModel)

fixedSlopeInterceptModel <- lm(loght ~ sTemp * Family, data = plantHeight)
randomSlopeModel <- lmer(loght ~ sTemp + (sTemp | Family), data = plantHeight)

summary(fixedSlopeInterceptModel)
summary(randomSlopeModel)
```

So, for the model with a different intercept per Family, the p-value decreases, and for intercept + interaction, it even becomes n.s. (although it should be noted that we get a different p-values for each family here).

The reason for the generally lower p-values (= higher power) of the mixed model is the lower flexibility of the mixed model:

For the fixed effect models, we lose 67 respectively 2x67 df if we add family as a predictor (68 families: 67 extra df for intercepts, 67 more for slopes). For the mixed model, it depends on the variance estimate: wide sd of the RE, lose nearly 67 df, narrow sd = lose nothing except the sd estimate -\> flexibility of a random effect model is adaptive (not fixed).

Follow-up: have a look at the variance estimates for the REs - how wide are they? Are the RE estimates effectively "free" to move?
:::

### Predictions {#sec-re-prediction}

Related to the interpretation of an RE is the question whether the REs should be included when you make predictions or calculate other outputs of the model, e.g. residuals.

- Unconditional predictions = predictions without the RE, i.e. the population-level curve from the fixed effects (see @tip-re-predterms)

- Conditional predictions = predictions with RE

In some, but unfortunately not all packages, the predict function can be adjusted. In lme4, this is via the option re.form, which is available in the predict and simulate function. Let look again at a simple random intercept model

```{r}
m = lmer(normexam ~ standLRT + sex +  (1 | school), data = Exam)
```

To generate predictions from this model, use the following syntax.

```{r, eval = F}
predict(m, re.form = NULL) # include all random effects
predict(m, re.form = ~0) # include no random effects, identical re.form = NA
predict(m, re.form = ~ (1|school)) # condition ONLY on 1|school
```

The third option is in this case identical to the first, but if we would add more REs to the model, it would differ.

::: callout-tip
Do you know how to get the correct help function for an S3 object? Concretely, how do we get the help for predicting with a fitted mixed model? `?predict` alone will not work, because this will call the general help. What you need is predict.CLASSNAME, where CLASSNAME is the class of the object on which you want to work on. In R, you get the class via

```{r}
class(m)
```

So, a model fitted with lme4 is of class lmerMod. If we want to get the help for predicting from such a class, we have to type

```{r}
?predict.merMod
```
:::

A second issue with predictions is their uncertainty. Because of the degrees of freedom problem (@sec-re-df), lme4 calculated no standard errors on predictions for a long time; since version 2.0, `predict(..., se.fit = TRUE)` gives approximate standard errors (with a warning). The more reliable option is the parametric bootstrap, which is implemented in lme4

For this, we generate first a function that generates predictions

```{r}
pred <- function(m) predict(m,re.form = NA)
```

Note regarding this function that

- We could add the argument newdata if we wanted to predict to newdata

- Here, we have re.form = NA, which is identical to re.form = \~0, meaning that we make unconditional predictions, i.e. the population-level curve from the fixed effects

Second, we now bootstrap this prediction to generate an uncertainty of the prediction

```{r, eval = F}
boot <- bootMer(m, pred, nsim = 100, re.form = NA)

# visualize results
plot(boot)

# get 95% confidence interval 
confint(boot)
```

### Model selection and hypothesis tests with mixed models {#sec-re-selection}

For model selection, the degrees of freedom problem (@sec-re-df) means that normal model selection techniques such as AIC or naive LRTs don't work on the random effect structure, because they don't count the correct degrees of freedom.

#### Selecting on fixed effects {#sec-re-fixed}

The easy part is the selection on the fixed effect structure - if we can assume that by changing the fixed effect structure, the flexibility of the REs doesn't change a lot (you would see this by looking at the RE sd), we can use all standard model selection methods (with all their caveats) as before on the fixed effect structure. All I will say about model selection on standard models applies also here: good for predictions, rarely a good idea if your goal is causal inference (see @sec-select-causality and the chapter on causal inference, @sec-causal).

::: callout-caution
#### Careful when selecting on models fit with REML

As discussed in @sec-re-reml, LMMs are per default fit with REML, whose likelihoods are not comparable between models with different fixed effects. When comparing models with different fixed effects via LRTs or AIC, refit the models with ML. See also [@verbyla2019]
:::

#### A priori selection on random effects {#sec-re-randomsel}

When changing the RE structure, you should not use standard AIC or LRTs, because of the df problem. For that reason, I would generally recommend to fix the RE structure *a priori* and keep selection on REs to a minimum.

What structure you should choose may depend on the experimental setting and scientific / biological expectations, but a base recipe for a limited number of data is:

1.  add random intercept on all obvious grouping variables
2.  check residuals per group (e.g. with the plot function in @sec-re-pergroup), or simulated LRT (see @sec-re-simlrt) and add random slope if needed

```{r, eval = F}
m1 <- lmer(y ~ x + (1|group))

plot(m1, 
     resid(., scaled=TRUE) ~ fitted(.) | group, 
     abline = 0)
```

I would deviate from this recipe and also add random slopes *a priori* when I either have sufficient power to support the random slopes without problems, or have good *a priori* reasons to expect random slopes.

#### Simulated LRTs {#sec-re-simlrt}

If you want to formally select on the variance structure, the best way is to use a simulated likelihood-ratio test (LRT), based on the parametric bootstrap, which we introduced in @sec-anova-test. In such a test, we make the same assumptions as in a standard LRT, which is:

1.  We want to compare 2 nested models M~0~, M~1~
2.  H~0~: simpler model M~0~ is correct
3.  Test statistic: M~1~/M~0~ or log(M~1~/M~0~) = log(M~1~) - log(M~0~)

However, unlike in the standard LRT, where we assumed that the test statistic is Chi^2^ distributed, we use the parametric bootstrap (see @sec-lm-bootstrap-param) to generate new data under H~0~, fit M~0~ and M~1~ to this simulated data, and thus generate a null distribution of the LR for these two models.

Simulated LRTs for mixed models are implemented in a number of packages. To my knowledge, one of the most general functions is DHARMa::simulateLRT. Here an example, in which we test if a random intercept is needed. Note that we fit the mixed model with ML (`REML = FALSE`), because its likelihood is compared to the likelihood of a model without random effects (see @sec-re-reml):

```{r, warning=F, message=F}
set.seed(123)
dat <- DHARMa::createData(sampleSize = 200, randomEffectVariance = 1, family = gaussian())

library(lme4)
m1 = lmer(observedResponse ~ Environment1 + (1|group), data = dat, REML = FALSE)
m0 = lm(observedResponse ~ Environment1 , data = dat)

DHARMa::simulateLRT(m0, m1)
```

If you want to understand the method in more detail, you can expand the hand-coded example of a simulated LRT below:

::: {.callout-tip collapse="true"}
#### Hand-coded simulated LRT

We use the models from the previous example - first, let's look at the observed LR

```{r}
observedLR = logLik(m1) - logLik(m0)
```

The log LR of m1 (the RE model) over m0 is 45, meaning that seeing the observed data under m1 is exp(45) times more likely than under m0.

As noted, the complex model will always fit better, so it is expected that the data is more likely under M1, but is the increase significant? In this case, the LR is actually so large that we actually wouldn't need to test. A rough rule of thumb is that you need a log LR of 2 for each df that you add, and here we have an RE with 10 groups, so even if we counted 1 df for each RE group, this should be easily significant. Nevertheless, let's do the formal test, and this time by hand.

First, we write a function that creates new LRs under H~0~, by simulating new data from the fitted M~0~, and re-fitting M~0~/M~1~ to this data, and return the LR.

```{r, warning=F, message=F}
resampledParameters = function(){
  newData = dat
  newData$observedResponse = unlist(simulate(m0))
  mNew0 = lm(observedResponse ~ Environment1, data = newData)
  mNew1 = lmer(observedResponse ~ Environment1 + (1|group), data = newData, REML = FALSE)
  return(logLik(mNew1) - logLik(mNew0))
}
```

Now, let's generate the null distribution. There will be a lot of warnings which you can usually ignore. The reason is that we simulate the data under H~0~, i.e. with no RE, and thus the mixed model estimates the RE variance to zero, which creates a warning.

```{r, warning=F, message=F}
nullDistribution = replicate(500, resampledParameters())
```

If we plot the null distribution, we see that if the data would really not have an RE, we would expect an increase of likelihood for the more complicated model of not more than 4 or so.

```{r}
hist(nullDistribution, breaks = 50, main = "Null distribution log(L(M1) / L(M0))")
```

However, what we actually observe is an increase of 45. I rescaled the x axis to make this visible.

```{r}
hist(nullDistribution, breaks = 50, main = "Null distribution log(L(M1) / L(M0))", xlim = c(-5,50))
abline(v = observedLR, col = "red")
```

No simulated LR exceeded the observed one, so p < 1/501 (about 0.002); a Monte Carlo p-value is never exactly 0.

```{r}
mean(nullDistribution > observedLR)
```
:::

### Variance partitioning / ANOVA {#sec-re-anova}

Also variance partitioning in mixed models is a bit tricky, as (see the type I/II/III ANOVA discussion in @sec-anova-collinearity) fixed and random components of the model are in some way "correlated".

First of all, note that you have to be careful because as most S3 functions in R, the anova function is overloaded, so depending on the class of the model object, different functions will be used.

Let's look at this for the case of a lmer model.

```{r, message=F, warning=F}
library(faraway)
library(lme4)
detach("package:lmerTest", unload=TRUE)
data(irrigation)
fit <- lmer(yield ~ irrigation + variety + (1|field), data = irrigation)
summary(fit)
```

If you run the anova command on this, it will call anova.merMod from package lme4, which runs a sequential (type I) ANOVA.

```{r}
anova(fit)
```

See discussions about the exact calculations of this function [here](https://stats.stackexchange.com/questions/71914/what-does-the-anova-command-do-with-a-lmer-model-object).

If you load lmerTest, the anova command will be overwritten by anova.lmerTest, which performs, as displayed in the output, a type III Anova with df estimated by Satterthwaite's method.

```{r, message=F, warning=F}
library(lmerTest)
fit <- lmer(yield ~ irrigation + variety + (1|field), data = irrigation)
anova(fit)
```

Arguably, this method is preferable over the base lme4 method if you are interested in the p-values of the model terms.

::: {#cau-re-typeiii .callout-caution}
#### Type III ANOVA for mixed models: coding and centering

The considerations from the chapter on ANOVA about contrasts and centering (@sec-anova-contrast, @cau-anova-typeiii) also apply to mixed models, with one difference between packages: the type III ANOVA of `lmerTest` does not depend on how factors are coded (it constructs the type III hypotheses independently of the contrasts used), but `car::Anova(fit, type = "III")` does, exactly as for an lm. Both still depend on centering: if a numeric predictor is part of an interaction, the main effects of the other variables are tested at the value 0 of this predictor. `emmeans::joint_tests()` also works for mixed models and avoids both problems. When reporting, state which method was used for the degrees of freedom (Satterthwaite as above, or Kenward-Roger).

Note also that in mixed models, the coding determines what random slopes for factors mean. If you use uncorrelated random effects (the `||` syntax), the coding even changes the fitted model, as the assumption of zero correlations depends on how the factor is coded.
:::

Another common question in ANOVA for mixed models is how much R2 is explained by fixed vs. random effects. Philosophically, the main point of discussion here is: Do you want to count the random effect variance as "explained", or "residual". In any case, a common approach is the marginal and conditional R2 proposed by @nakagawa2013general and @nakagawa2017coefficient, which is implemented in the `MuMIn` package. With this, we can run

```{r chunk_chapter4_chunk33, echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
library(MuMIn)
r.squaredGLMM(fit) 
```

Interpretation

- **R^2^m**: Marginal R^2^ value associated with fixed effects (the name is historical; in the terminology of @tip-re-predterms, it is based on the unconditional predictions).
- **R^2^c**: Conditional R^2^ value associated with fixed effects plus the random effects.

It is important to note that this is a description of the fitted model, not necessarily valid inference - the reason is that if you add a lot of REs, you will tend to get a lot of variance explained by the RE structure. If you want to correct for this, one option is to calculate R^2^ values under a cross-validation.

### Variance partition coefficients (VPCs) {#sec-re-vpc}

An interesting variation of using mixed models is to use them as an alternative to ANOVA to understand how much variation in a response variable is explained by different factor variables (for applications in ecology, see [@leite2024]). Consider the following example of simulated data, where I created 3 categorical variables with an increasing number of levels per variable, and essentially a "normal" distribution of effects for the different levels of those. In other words, these are exactly the assumptions of a mixed model, where we have 4 normally distributed errors: 3 for the 3 groups (with different group sizes) and 1 residual error, all with the same sd of 1.

```{r}
set.seed(123)

k1 = 4
k2 = 8
k3 = 20
rep = 5

RE1 = rnorm(k1)
RE2 = rnorm(k2)
RE3 = rnorm(k3)

dat = expand.grid(1:k1,1:k2,1:k3, 1:rep)
dat$y = RE1[dat$Var1] + RE2[dat$Var2] + RE3[dat$Var3] + rnorm(nrow(dat))
dat[1:4] <- lapply(dat[1:4], as.factor)
```

To see how much variation is created by each of the factors, the standard way would be to fit an ANOVA (see the chapter on ANOVA, @sec-anova, if you don't recall the details). As the design is perfectly balanced and orthogonal, it doesn't matter if we fit type I,II,III ANOVA:

```{r}
summary(aov(y ~ Var1+Var2+Var3, data = dat))
# car::Anova(lm(y ~ Var1+Var2+Var3, data = dat)) # identical
```

The alternative would be to fit a random effect model with the 3 grouping variables as random intercepts

```{r}
library(lme4)
library(lmerTest)
fit <- lmer(y ~ (1|Var1) + (1|Var2) + (1|Var3), data = dat)
```

After having fit the model, we can look at the random effect table

```{r}
x = VarCorr(fit)
x
```

Which pretty much already looks like an ANOVA table. If you standardize the RE variances to sum of all variance, you obtain the so-called **variance partition coefficients (VPCs).**, which are often seen as an alternative to ANOVA

```{r}
x = data.frame(x)
x$vcov / sum(x$vcov)
```

::: callout-important
Note that the results of a VPC can be very different from an ANOVA. In an ANOVA, you look at the variance explained in the response. In VPCs, you compare the variance between different groups in the data. This will tend to be similar to ANOVA if observations are balanced in groups, but it can be very different if they are not. In this behavior, VPCs resemble standardized effect sizes (which can also differ in ranking variables from ANOVA if the distribution of the predictor is very skewed). I would therefore recommend to interpret VPCs as "effect sizes of categorical predictors", rather than as ANOVA shares.
:::

We can get (note: approximate) p-values on the terms by running a random effect anova

```{r}
lmerTest::ranova(fit)
```

The quality of the p-value approximation is somewhat unclear. Alternatively, you could run parametric-bootstrap (simulated) LRTs (see @sec-re-simlrt). In a frequentist framework, it will be difficult to obtain CIs on the VPCs, unless you bootstrap. Alternatively, you could run this in a Bayesian framework via brms.

As a final step, let's test out if we test for factors that did not have an effect in the model. In this case, I create three two-way interactions as well as another RE on the factor replicate, which did not have an effect on the response

```{r, warning=F, message = F}
dat$Var12 = factor(paste(dat$Var1,dat$Var2))
dat$Var23 = factor(paste(dat$Var2,dat$Var3))
dat$Var13 = factor(paste(dat$Var1,dat$Var3))

summary(aov(y ~ (Var1+Var2+Var3)^2 +Var4, data = dat))

fit <- lmer(y ~ (1|Var1) + (1|Var2) + (1|Var3) + (1|Var12) + (1|Var23) + (1|Var13) +  (1|Var4)  , data = dat)
VarCorr(fit)
lmerTest::ranova(fit)
```

In this case, both methods attribute some variation to the null terms, but also both methods realize that these terms are (mostly) n.s., which is reassuring! The term Var12 is around 0.05, significant in ANOVA and n.s. in VPC, suggesting maybe that the VPC could be a bit more conservative, although I suspect that this is highly context and package dependent (as different packages fit the REs differently). Note that I have suppressed boundary / convergence warnings in the fit of the mixed model, which should be of course solved in a real application.

::: callout-tip
#### VPC vs. ANOVA

So, when and why should you prefer VPC over ANOVA?

A considerable advantage of the VPCs over ANOVA is that they compute orders of magnitudes faster than a standard anova in large datasets with many group levels.

Disadvantages are measures of uncertainty (p-values, CIs) may be problematic to obtain. Moreover, you should also note that the estimation of the VPCs depends on the ability of the mixed models to estimate the correct RE variance. To obtain unbiased estimates of the RE variance is notoriously difficult, as evidenced by the different methods to estimate the RE variances in different packages, as well as the REML / ML estimation debate (see @sec-re-reml). The issues are usually not a primary concern, because we usually do not interpret the RE variances directly, but they are only indirectly used to account for the group level effects. In the VPC analysis, however, they are directly used, and thus particular care should be taken, e.g., in checking if RE distributions are indeed normal etc. (see, e.g., @sec-re-normality).
:::

## Take-home messages {#sec-re-takehome}

::: callout-tip

1.  Model a grouping variable as a random effect if you are not interested in its individual estimates: it accounts for the correlation within groups and uses fewer df than a fixed effect. Use a fixed effect if you need unbiased group estimates, because random effect estimates are shrunk towards the grand mean.
2.  Fix the random effect structure a priori (random intercepts for all grouping variables; random slopes if expected or if the residuals per group suggest them). Do not select it with AIC or standard LRTs, because the df of a random effect are unclear; use a simulated LRT if you need a test.
3.  Check that group IDs are unique across higher-level groups, or use the nested syntax `(1 | group / subgroup)`; otherwise different groups are merged.
4.  Fit with REML, but refit with ML when you compare models with different fixed effects by LRT or AIC. `lmer()` reports no p-values because of the df problem; use `lmerTest` (approximate df) or the parametric bootstrap.
5.  Check the residuals as for an lm, plus convergence (center and scale predictors; a singular fit often means a variance estimated at zero; for grouping variables with few levels, see @sec-re-usage) and the random effects (normality, correlation with predictors). State whether predictions and residuals are conditional or unconditional, and use the parametric bootstrap for confidence intervals of predictions.
:::

## Case studies {#sec-re-cases}

::: {#case-studies}
:::

### Honeybee data {#sec-re-honeybee}

We use a dataset on bee colonies infected by the American Foulbrood (AFB) disease. The study measured spores of the bacterium "Paenibacillus larvae" (variable Spobee) that infect honeybee larvae, as well as the degree of the infection of the colony (integer: 0 (none), 1 (minor), 2 (moderate), 3 (major)).

The scientific question is if colonies with a higher degree of infection also exhibit more spores, i.e. if infection predicts spores. Additional variables that you can consider are BeesN (Bees per hive) and Hive (ID of the respective hive. (integer, 1 - 24).)

::: {.callout-caution icon="false"}
#### Exercise

Fit an appropriate linear mixed model on this data.

```{r chunk_chapter4_task_3, message=FALSE, warning=FALSE, eval = TRUE, purl=FALSE}
library(EcoData)
str(bees)
```

Perform the data analysis, according to the hypothesis discussed in the course.
:::

::: {.callout-tip collapse="true" appearance="minimal" icon="false"}
#### Solution

- Because of the skewed distribution of Spobee, it is better to use log(Spobee + 1) or something similar as response (else you will not get nice residuals)
- I decided to add BeesN in the regression because it could be a confounder (more on confounding in the chapter on causal inference, @sec-causal)
- It is natural to model hive as a random intercept (1\|Hive). I don't expect that the infection \~ spores relationship should be different between hives, but can test for this after fitting the model

```{r chunk_chapter4_task_4, message=FALSE, warning=FALSE}
library(lme4)
fit <- lmer(log(Spobee + 1) ~ Infection + BeesN + (1|Hive), data = bees)
summary(fit)
```

Standard residual plots look fine, possible minor evidence for heteroskedasticity

```{r}
plot(fit)
```

Let's have a look at the random effects

```{r}
x = ranef(fit)
qqnorm(x$Hive$`(Intercept)`)
qqline(x$Hive$`(Intercept)`)
shapiro.test(x$Hive$`(Intercept)`)
```

Looks good. Could also check with DHARMa. 

```{r}
simulateResiduals(fit, plot = T)
```

It doesn't make sense to add a random slope here: Infection is constant within each hive, so a random slope for Infection by Hive cannot be estimated.

```{r}
plot(fit, 
     resid(., scaled=TRUE) ~ fitted(.) | Hive, 
     abline = 0)
```
:::

### College student performance {#sec-re-college}

The GPA (college grade point average) data is a **longitudinal** data set (also named **panel data**, German: "Längsschnittstudie". A study repeated at several different moments in time, compared to a **cross-sectional study** (German: "Querschnittstudie") which has several participants at *one time*). In this data set, 200 college students and their GPA have been followed 6 consecutive semesters. Look at the GPA data set, which can be found in the `EcoData`{.R} package:

```{r chunk_chapter4_chunk29, echo=TRUE, eval=F}
library(EcoData)
str(gpa)
```

In this data set, there are GPA measures on 6 consecutive **occasions**, with a **job** status variable (how many hours worked) for the same 6 occasions. There are two student-level explanatory variables: The **sex** (1 = male, 2 = female) and the high school **gpa**. There is also a dichotomous student-level outcome variable, which indicates whether a student has been **admitted** to the university of their choice. Since not every student applies to a university, this variable has many missing values. Each **student** and each **year** of observation have an id.

::: {.callout-caution icon="false"}
#### Exercise

Analyze if the student's GPA improves over time (**occasion**)! Here a few hints to look at:

- Consider which fixed effect structure you want to fit. For example, you might be interested if males and females differ in their temporal trend
- Student is the grouping variable -\> RE. Which RE structure do you want to fit? A residual plot may help
- For your benefit, have a look at the difference in the regression table (confidence intervals, coefficients and p-values) of mixed and corresponding fixed effects model. You can also look at the estimates of the mixed effects model (hint: `?ranef`{.R}).
- After having specified the mixed model, have a look at residuals.
- Which predictors vary between students, and which within students over time? What does this mean for the random effect structure (see @tip-re-design)?
:::

::: {.callout-tip collapse="true" appearance="minimal" icon="false"}
#### Solution

**1. Design.** Sex and the high school GPA are constant within a student, i.e. they are student-level predictors, replicated only 200 times. Occasion and job vary within students. The 6 measurements of a student are not independent, so we need (at least) a random intercept for student (see @tip-re-design). In general: (1) for each predictor, decide whether it varies between or within the grouping units; (2) add a random intercept for every grouping unit; (3) add random slopes for within-unit predictors whose effect plausibly differs between units; (4) check the residuals for correlation over time and for changing variance.

```{r chunk_chapter4_task_2, message=FALSE}
library(lme4)
library(EcoData)
gpa$nJob = as.numeric(gpa$job) - 1 # hours worked per week (0 to 4), assuming equal distances between categories
table(table(gpa$student))      # 200 students with 6 measurements each
```

**2. Why the random intercept is needed.** Let's compare a model that ignores the grouping with the mixed model:

```{r}
fitLM = lm(gpa ~ occasion * sex + nJob, data = gpa)
fitRI = lmer(gpa ~ occasion * sex + nJob + (1|student), data = gpa)
summary(fitLM)$coefficients[, 1:2]
summary(fitRI)$coefficients[, 1:2]
```

The standard error of sex, a student-level predictor, is too small in the lm, because the lm treats the 6 measurements per student as independent observations (pseudoreplication). For occasion, which varies within students, the mixed model is even more precise than the lm, because each student is compared with themselves (the occasion effect is estimated within students). Also the estimate for job changes: in the lm, it partly reflects differences between students (students who work more have lower grades overall).

**3. How to code time.** If we fit occasion as a factor, we see how the mean GPA changes from semester to semester:

```{r}
fitF = lmer(gpa ~ factor(occasion) + sex + nJob + (1|student), data = gpa)
fixef(fitF)
```

The increase per semester is roughly constant, so a linear effect of occasion is adequate (and needs fewer parameters).

**4. Random slope for time.** It is plausible that students differ in their trends over time. The residuals of the random intercept model for the first 20 students confirm this:

```{r}
lattice::xyplot(residuals(fitRI) ~ occasion | student, data = gpa,
                subset = as.numeric(student) <= 20, type = c("p", "r"))
```

So we add a random slope for occasion (if you want a formal test, use a simulated LRT, see @sec-re-simlrt):

```{r}
fitRS = lmer(gpa ~ occasion * sex + nJob + (occasion|student), data = gpa)
summary(fitRS)$coefficients[, 1:2]
```

Note that the standard error of occasion increases compared to the random intercept model: if students differ in their trends, a model with a random intercept only is overconfident about the average trend (see @sec-re-usage).

Steps 5 and 6 use tools from the chapters on correlation structures (@sec-corr) and dispersion models (@sec-disp); read them as a preview of how the checks continue. Both matter, because correlation over time and changing variance can make standard errors and p-values wrong.

**5. Is a temporal correlation structure needed?** The measurements of a student are a short time series. We can add an AR1 structure within students (see @sec-corr-temporal) and compare the models:

```{r, message=FALSE}
library(glmmTMB)
gpa$fOccasion = factor(gpa$occasion)
fitRS2 = glmmTMB(gpa ~ occasion * sex + nJob + (occasion|student), data = gpa)
# AR1: correlation between neighboring semesters (time must be a factor)
fitAR = glmmTMB(gpa ~ occasion * sex + nJob + (occasion|student) +
                  ar1(fOccasion + 0 | student), data = gpa)
AIC(fitRS2, fitAR)
```

The AR1 structure lowers the AIC by about 4, but glmmTMB warns that the fit did not converge properly (singular convergence), and the estimates are degenerate (look at the random effect variances and their correlation in `summary(fitAR)`). With only 6 time points per student, the random intercept and slope already capture most of the temporal dependence, and the AR1 parameter is hard to estimate on top of them. We therefore keep the simpler model; this decision rests on the fit problems and the short series, not on the AIC.

**6. Heteroskedasticity.** The residual spread decreases in the later semesters:

```{r}
boxplot(residuals(fitRS) ~ gpa$occasion, xlab = "occasion", ylab = "residuals")
```

How to model this is explained in the chapter on dispersion models (@sec-disp). As a teaser, glmmTMB allows a formula for the dispersion; `dispformula = ~ occasion` means that the residual standard deviation may change with occasion:

```{r, message=FALSE}
fitDisp = glmmTMB(gpa ~ occasion * sex + nJob + (occasion|student),
                  dispformula = ~ occasion, data = gpa)
summary(fitDisp)
AIC(fitRS2, fitDisp)

library(DHARMa)
plot(simulateResiduals(fitDisp))
```

The variable dispersion model is strongly supported by the AIC (this compares variance structures, not random effects), and the residuals are much better, though the quantile tests still show slight deviations. The answer to the question: GPA increases over time, by about 0.09 points per semester for male students, and somewhat more for female students (see the `occasion:sexfemale` interaction).
:::
