```{r, include=FALSE}
set.seed(42)
```

# Dispersion models {#sec-disp}

::: callout-note
#### In this chapter

In this chapter, you will learn

-   why heteroskedasticity makes p-values of the linear model and ANOVA unreliable, and how to fix it by a transformation or by modeling the variance with `gls` or `glmmTMB` (@sec-disp-lm);
-   how to model a variable dispersion in GLMMs (@sec-disp-glmm);
-   what to do about outliers and non-normal residuals: robust and quantile regression (@sec-disp-outlier).

**Workflow steps** (@sec-intro-workflow): step 5 (residual checks: changing variance).

Prerequisites: the linear model and its residual checks (@sec-lm), GLMMs (@sec-glm) for the second part.
:::

Modeling dispersion means that we describe how the expected variance changes as a function of the mean or predictor variables. The residual problem that we address by that is known as heteroskedasticity (see @sec-lm-mean-variance and @sec-lm-residual-lm).

## Heteroskedasticity in the linear model {#sec-disp-lm}

::: {#heteroskedasticity-in-the-linear-model}
:::

Heteroskedasticity means that the residual variance is not constant, for example because it depends on a predictor or on the mean. The easiest way to understand it is in the linear model. Let's look at an extreme example that we create by simulating some data

```{r chunk_chapter4_chunk8, echo=TRUE, eval=TRUE}
set.seed(125)

data = data.frame(treatment = factor(rep(c("A", "B", "C"), each = 15)))
data$observation = c(7, 2 ,4)[as.numeric(data$treatment)] +
  rnorm( length(data$treatment), sd = as.numeric(data$treatment)^2 )
boxplot(observation ~ treatment, data = data)
```

Especially p-values and confidence intervals of `lm()`{.R} and ANOVA can react quite strongly to such differences in residual variation. So, running a standard `lm()`{.R} / ANOVA on this data is not a good idea - in this case, we see that no regression effect is significant, and neither is the ANOVA (p = 0.27), although the true group means differ (7, 2, 4, we simulated them). The unequal residual variances distort the standard errors and p-values (here they cost us power); let's see what happens when we model the variances.

```{r chunk_chapter4_chunk9, echo=TRUE, eval=TRUE}
fit = lm(observation ~ treatment, data = data)
summary(fit)
summary(aov(fit))
```

::: callout-note
A subtlety about heteroskedasticity in the normal model is that it especially affects p-values when the residual variance changes with the predictor itself, as illustrated by the simulation below (no true effect, so p-values should be uniform)

```{r, fig.height=3.5}
par(mfrow = c(1, 2))
# error sd is variable, but uncorrelated with predictor

getP <- function(effect = 0, n = 1000){
  sd = rgamma(n, shape = 2)
  #hist(sd)
  x = runif(n)
  y = effect * x + rnorm(n, sd = sd)  
  fit <- lm(y ~ x)
  table <- summary(fit)
  return(table$coefficients[2,4])
}

out = replicate(1000, getP())
hist(out, breaks = 20, main = "variance independent of x")


# error sd is variable and correlated with predictor (identical in this case)

getP2 <- function(effect = 0, n = 1000){
  sd = rgamma(n, shape = 2)
  y = effect * sd + rnorm(n, sd = sd)  
  fit <- lm(y ~ sd)
  table <- summary(fit)
  return(table$coefficients[2,4])
}

out2 = replicate(1000, getP2())
hist(out2, breaks = 20, main = "variance depends on x")
```

Left: the p-values are uniform under H0, i.e. calibrated (about 5% false positives at alpha = 0.05). Right: strong excess of small p-values (here about 30% false positives): inflated type I error.
:::

So, what can we do?

### Transformation {#sec-disp-transform}

One option is to search for a transformation of the response that improves the problem - If heteroskedasticity correlates with the mean value, one can typically decrease it by some sqrt or log transformation, but this is often difficult, because this may also conflict with keeping the distribution normal.

### Modeling the variance / dispersion {#sec-disp-model}

The second, more general option is to model the variance, i.e. to fit a model in which the variance is not fixed. We will discuss two packages in R that allow you to model the dispersion.

The first (traditional) option is to use `nlme::gls`{.R}. GLS = *Generalized Least Squares* (not to be confused with GLMs, @sec-glm: GLS keeps the normal distribution, but allows non-constant variance and correlated errors). In the `gls` function, you can specify a dependency of the residual variance on a predictor or the response via the `weights` argument. There are different types of dependencies that you can specify, see `?varFunc`{.R}. In our case, we will use the `varIdent` function, which allows you to specify a different variance per treatment.

```{r chunk_chapter4_chunk10, echo=TRUE, eval=TRUE}
library(nlme)

fit = gls(observation ~ treatment, data = data, 
          weights = varIdent(form = ~ 1 | treatment))
summary(fit)
```

If we check the ANOVA, we see that, unlike before, we get a significant effect of the treatment

```{r chunk_chapter4_chunk11, echo=TRUE, eval=TRUE}
anova(fit)
```

The second option for modeling variances is to use the `glmmTMB` package. Here, you can specify an extra regression formula for the dispersion. For the Gaussian family, this formula models the residual standard deviation on the log scale. If we fit this:

```{r chunk_chapter4_chunk12, echo=TRUE, eval=TRUE}
library(glmmTMB)

fit = glmmTMB(observation ~ treatment, data = data, 
              dispformula = ~ treatment)
summary(fit)
```

We get 2 regression tables as outputs - one for the effects, and one for the dispersion. We see, as expected, that the dispersion is higher in groups B and C compared to A: for the Gaussian family the dispersion formula models the residual standard deviation on the log scale, so exp(1.55) = 4.7 means that the SD in B is about 4.7 times that in A. `gls` reports the same quantities as SD ratios relative to A (4.7 for B and 11.8 for C, with confidence intervals from `intervals()` on the gls fit), whereas `glmmTMB` additionally gives z-tests and p-values for these differences.

::: {.callout-caution icon="false"}
#### Exercise — Variance modeling

Take this plot of Ozone \~ Solar.R using the airquality data. Clearly there is heteroskedasticity in the relationship:

```{r chunk_chapter4_chunk14, echo=TRUE, eval=TRUE}
plot(Ozone ~ Solar.R, data = airquality)
```

We can also see this when we fit the regression model:

```{r chunk_chapter4_chunk15, echo=TRUE, eval=TRUE}
m1 = lm(Ozone ~ Solar.R, data = airquality)
par(mfrow = c(2, 2))
plot(m1)
```

We could of course consider other predictors, but let's say we want to fit this model specifically

1.  Try to get the variance stable with a transformation.
2.  Use the `gls` function (package `nlme`) with the untransformed response to make the variance dependent on Solar.R. Hint: Read in `varClasses` and decide how to model this.
3.  Use `glmmTMB` to model heteroskedasticity.
:::

::: {.callout-tip collapse="true" appearance="minimal" icon="false"}
#### Solution

1.  **Transformation**

```{r}
m3 = lm(sqrt(Ozone) ~ Solar.R, data = airquality)
```

2.  **GLS**

```{r}
m3 = nlme::gls(Ozone ~ Solar.R, weights = varPower(form = ~Solar.R), data = airquality[complete.cases(airquality),])
```

3.  **glmmTMB**

```{r}
m4 = glmmTMB(Ozone ~ Solar.R, dispformula = ~Solar.R, data = airquality)
```
:::

## Heteroskedasticity in GLMMs {#sec-disp-glmm}

::: {#heteroskedasticity-in-glmms}
:::

GLM(M)s can be heteroskedastic as well, i.e. dispersion depends on some predictors. This is the same idea as the dispersion problems in GLMs (@sec-glm-dispersion): the variance is not what the distribution assumes, but here we model it explicitly. In `glmmTMB`, you can make the dispersion of the negative Binomial dependent on a formula via the `dispformula` argument, analogous to the variance modeling with `gls` and `glmmTMB` for the linear model (@sec-disp-model).

Variance problems would show up when plotting residuals against predicted and predictors (see @sec-glm-dharma). In the section on recognizing overdispersion (@sec-glm-overdispersion-detect), we saw some variance problems in the Salamander model. We could add a variable dispersion model via

```{r chunk_chapter5_chunk21, echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
m3 = glmmTMB(count ~ spp + mined + (1|site), family = nbinom1,
             dispformula = ~ spp + mined ,  data = Salamanders)
summary(m3)
library(DHARMa)
res = simulateResiduals(m3, plot = T)

par(mfrow = c(1, 2))
plotResiduals(res, Salamanders$spp)
plotResiduals(res, Salamanders$mined)
```

## Excursion: outliers, robust and quantile regression {#sec-disp-outlier}

::: {#excursion-outliers-robust-and-quantile-regression}
:::

What can we do if, after accounting for the functional form (@sec-lm-functional-form), response transformation (@sec-disp-transform) and variance modeling (@sec-disp-model), the Q-Q plot of the residuals (@sec-lm-residual-lm) still shows non-normality, in particular strong outliers? Here simulated example data with strong outliers / deviations from normality:

```{r chunk_chapter4_chunk16, echo=TRUE, eval=TRUE}
set.seed(123)

n = 100
concentration = runif(n, -1, 1)
growth = 2 * concentration + rnorm(n, sd = 0.5) +
  rbinom(n, 1, 0.05) * rnorm(n, mean = 6*concentration, sd = 6)
plot(growth ~ concentration)
```

Fitting the model, we see that the distribution is too wide:

```{r chunk_chapter4_chunk17, echo=TRUE, eval=TRUE}
fit = lm(growth ~ concentration)
par(mfrow = c(2, 2))
plot(fit)
```

What can we do to deal with such distributional problems and outliers?

- **Removing** - Bad option, hard to defend, reviewers don't like this - if at all, better show robustness with and without outlier, but result is sometimes not robust.
- **Change the distribution** - Fit a model with a different distribution, i.e. GLM (@sec-glm) or other.
- **Robust regressions**.
- **Quantile regression** - A special type of regression that does not assume a particular residual distribution.

### Robust regression {#sec-disp-robust}

Robust methods generally refer to methods that are robust to violation of assumptions, e.g. outliers. More specifically, standard robust regressions typically downweight data points that have a too high influence on the fit. See <https://cran.r-project.org/web/views/Robust.html> for a list of robust packages in R.

```{r chunk_chapter4_chunk18, echo=TRUE, eval=TRUE}
# This is the classic method.
library(MASS)

fit = rlm(growth ~ concentration) 
summary(fit)
# No p-values and not sure if we can trust the confidence intervals.
# Would need to bootstrap by hand!

# This is another option that gives us p-values directly.
library(robustbase)

fit = lmrob(growth ~ concentration) 
summary(fit)
```

### Quantile regression {#sec-disp-quantile}

Quantile regressions don't fit a line with an error spreading around it, but try to fit a quantile (e.g. the 0.5 quantile, the median) regardless of the distribution. Thus, they do not assume a particular residual distribution or constant variance, but functional form and independence still matter, and the median is not the mean.

```{r chunk_chapter4_chunk19, echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
library(qgam)

dat = data.frame(growth = growth, concentration = concentration)

fit = qgam(growth ~ concentration, data = dat, qu = 0.5) 
summary(fit)
```

## Take-home messages {#sec-disp-takehome}

::: callout-tip

1.  Check the residuals against the predicted values and against each predictor. With heteroskedasticity, p-values and confidence intervals of `lm()` and ANOVA can be badly wrong, mainly for the predictor the variance is related to; in our example, a non-significant ANOVA became significant once the variance was modeled.
2.  A transformation (e.g. sqrt or log) can remove the problem, but may conflict with normality. The more general solution is to model the variance: `gls` with `varIdent` or `varPower`, or `glmmTMB` with `dispformula`. For the Gaussian family, `dispformula` models the residual standard deviation on the log scale.
3.  In GLMMs, the dispersion can also depend on predictors (`dispformula`, e.g. with the negative binomial). Check again with simulated residuals plotted against the predictors.
4.  Do not simply remove outliers; if you do, show the results with and without them. Better: change the distribution, or use robust or quantile regression.
5.  Quantile regression assumes no particular residual distribution, but the functional form and independence still matter, and a median regression describes the median, not the mean.
:::
